\documentclass{beamer}
\usetheme{metropolis}           % Use metropolis theme
\title{A strongly consistent index for email using git and MirageOS}
\date{}
\author{Oliver Hope}
\institute{Jesus College}
\begin{document}

  \maketitle

  \begin{frame}{What?}
    What has been built?

    \begin{itemize}
      \item A library called \emph{gitmaildir} - An OCaml library to interface with and perform operations on the type of maildir in a git store that I have created
      \item An application called \emph{gitmaildir\_cli} - A command line client to allow inteaction with a gitmaildir form the commandline
      \item An application called \emph{gitmaildir\_daemon} - A daemon that keeps a standard maildir in sync with a gitmaildir allowing applications that use standard maildirs to interact without needing further changes
    \end{itemize}
  \end{frame}

  \begin{frame}{Why?}
    Maildir is a well known mail storage format for unix-like systems. Its benefits are:
    \begin{itemize}
      \item Allows lots of concurrent access
      \item Mail client does not have to worry as managed by the filesystem
    \end{itemize}
    However it has some downsides:
    \begin{itemize}
      \item We can never ensure that the directory is in a consistent state
      \item This makes operations that rely on indexing more difficult
    \end{itemize}
  \end{frame}

  \begin{frame}{How?}
    It turns out that MirageOS has some libraries that are useful in working with both git stores and email parsing using OCaml. (ocaml-git and MrMime)

    Therefore the entire project is written in OCaml making heavy use of these libraries.

    Everything that modifies the gitmaildir is written using a library called Lwt. This is a form of promise and allows operations to be combined in a monadic style. Another benefit of this is that we can easily parallelize functionality when necessary.
  \end{frame}

  \begin{frame}{What's left?}
    While the core features are complete, this provides a basis to implement more interesting extensions including:
    \begin{itemize}
      \item Separate branches for mobile clients that strip data-heavy content and attachments from emails
      \item Branches per reply thread for a better insight and cleaner overview of long-running email threads
      \item Intelligent rollback in case of accidental deletion or modification
    \end{itemize}
  \end{frame}
\end{document}
